\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-11 docs/parco13/abstract.tex
Go through first two pages of parco paper

\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\sR}{\mathbb{R}}

%%%%%%%%%%% End of definitions
\begin{document}
\begin{frontmatter}          % The preamble begins here.
%
%\pretitle{}
\title{ForestClaw:
        Hybrid forest-of-octrees AMR for hyperbolic conservation laws}
\runningtitle{ForestClaw}
%\subtitle{}

@@

\begin{abstract}
%
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ of dimensions $m^d$, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to work with the p4est AMR library since it allows to compose the mesh
from multiple mapped octrees, thus enabling the cubed sphere and other
nontrivial multiblock geometries.
%
\end{abstract}

@@

\section*{Introduction}

@@

To enable hardware acceleration for parallel dynamic AMR, we like to build upon
the forest-of-octrees paradigm because of its low overhead and proven
scalability \cite{BursteddeWilcoxGhattas11}.  As an important extension, we go
beyond the traditional high-order element and define each mesh element to be a
dense compute patch with $m^d$ DOFs.  In fact, this approach resembles
block-structured AMR \cite{ColellaGravesKeenEtAl07}
\todo{favorite BoxLib citation}
except that the patches are not overlapping, which enables us to capitalize on
our previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  A particular implementation of such a
patch is available with the Clawpack software \cite{LeVeque97} that has been
designed to solve hyperbolic conservation laws on a uniform compute patch and
successfully used in the context of block-structured AMR \todo{AMRClaw
citation}.

@@

\section*{Design principles}

Now combine.

Experience from integration with large-scale adaptive-mesh PDE solvers.
Design goals: Lightweight, modular, reuse.

Separation of mesh on the one hand from discretization and solvers on the other.

Mesh information is discrete (tree nodes have integer coordinates in
$(0, 2^L($ where $L$ is the maximum allowed refinement level of the tree.

sorted ordering of tree nodes, lookup tables, .

rank as is common with parallelization using the MPI framework
%\cite{Forum94, SnirOttoHuss-LedermanEtAl96}.

Challenge: patch neighborhood information is known to the mesh backend
but needs to be used to handle ghost information.

\section*{Hybrid forest-of-octrees AMR}

@@

\section*{Patch-based numerics at the leaf level}

@@

\section*{Numerical results}

>>>>

\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}\xspace}
\newcommand{\sR}{\mathbb{R}}
\newcommand{\forestclaw}{Forestclaw\xspace}
\newcommand{\pforest}{\texttt{p4est}\xspace}

%%%%%%%%%%% End of definitions
\begin{document}
\begin{frontmatter}          % The preamble begins here.
%
%\pretitle{}
\title{\forestclaw:
        Hybrid forest-of-octrees AMR for hyperbolic conservation laws}
\runningtitle{\forestclaw}
%\subtitle{}

@@

\begin{abstract}
%
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ of dimensions $m^d$, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to work with the \pforest AMR library since it allows to compose the mesh
from multiple mapped octrees, thus enabling the cubed sphere and other
nontrivial multiblock geometries.
%
\end{abstract}

@@

\section{Introduction}

@@

To enable hardware acceleration for parallel dynamic AMR, we like to build upon
the forest-of-octrees paradigm because of its low overhead and proven
scalability \cite{BursteddeWilcoxGhattas11}.  This approach identifies each
octree leaf with a mesh element.  In this work, we go
beyond the traditional high-order element and define each element to be a
dense compute patch with $m^d$ DOFs.  In fact, this approach resembles
block-structured AMR \cite{ColellaGravesKeenEtAl07}
\todo{favorite BoxLib citation}
except that the patches are not overlapping, which enables us to capitalize on
our previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The Clawpack software \cite{LeVeque97}
provides a popular implementation of such a patch.  It has
been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR \todo{AMRClaw
citation}.

@@

\section{Design principles}

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the Clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{Clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
structure & forest of octrees & patch of $m^d$ DOFs \\
atomic unit & octree leaf & one DOF \\
parallelization & MPI & threads (Manyclaw variant) \\
memory access & distributed & shared \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python \\
%\hline
\end{tabular}
\end{center}
The proposed 1:1 correspondence between a leaf and a patch thus combines two
hithertwo disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified and performant codes.
% if the integration is done well
\item We preserve the separation of the mesh on one hand from the
discretization and solvers on the other.
\item The AMR metadata (24 bytes per leaf) is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid and often referred
to as MPI+X.
\end{enumerate}

% yields periodicity as a special case
A particular feature of \forestclaw is that the generic handling of multiblock
geometries is inherited from \pforest, identifying each octree as a block, and
each leaf as a patch.  Each block is understood as a reference unit cube with
its own geometric mapping.  The connectivity of the blocks is static and can be
created by external hexahedral mesh generators, elimintating the need to encode
it by hand.

A main challenge is presented by the fact that the patch neighborhood is only
known to the mesh backend, from which it needs to be propagated to the
numerical code in \forestclaw that implements the interaction with ghost
patches.  To this end, we define an interface that allows distributed read-only
access to the sequence of blocks, the list of patches for each, and a lookup of
neighbor patches (and their relative orientation which is nontrivial between
blocks).  The exchange of ghost data and parallel partitioning is provided
transparently by the backend.  Mesh modification directives, such as local
refinement and coarsening, are called from \forestclaw and relayed to the mesh
backend.

% AMR Metadata $\ll$ numerical data

% Experience from integration with large-scale adaptive-mesh PDE solvers.
% Design goals: Lightweight, modular, reuse.

% Mesh information is discrete (tree nodes have integer coordinates in
% $(0, 2^L($ where $L$ is the maximum allowed refinement level of the tree.

% Dimension independence

% rank as is common with parallelization using the MPI framework
% \cite{Forum94, SnirOttoHuss-LedermanEtAl96}.

\section{Parallelization}

@@

fast (O(log N/P)) neighbor and parent/child lookups
horizontal/vertical tree traversal

subcycling: level-wise load balancing as done in structured AMR or
geometric adaptive multigrid \cite{SundarBirosBursteddeEtAl12}

\todo{What is BearClaw (tree-based / S.\ Mitran)?}

\section{Patch-based numerics at the leaf level}

@@

\section{Numerical results}

\section*{Acknowledgements}

The authors acknowledge valuable discussion with Randy LeVeque, Marsha Berger,
and Hans-Petter Langtangen.  The leaf/patch paradigm was independently presented
by B.\ as part of a talk at the SCI Institute, Utah, in July 2011.
